---
title: "Data 607 Project 1"
author: "Kayleah Griffen"
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this project, a text file with chess tournament results was provided. The objective was to create an R Markdown file that generates a .CSV file with the following information for all of the players:

Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents 

For example:

Gary Hua, ON, 6.0, 1794, 1605

## Processing

Load required packages.

```{r, message=FALSE }
library(RCurl)
library(tidyverse)
```

Read in the text file from github and take a look at the start.

```{r get-text}
x <- getURL("https://raw.githubusercontent.com/klgriffen96/spring23_data607_proj1/main/chess_scores.txt")
glimpse(x)
```

Text files are generally broken into new lines using `\r\n` which can be seen in this file. Start by splitting on `\r\n` so each line is separated.

```{r split-1}
# Split x 
s <- str_split_fixed(x, "\r\n", n=Inf)
# View first ten entries
s[1:10]
```
From the file extract the:

- Starting row
- Last player ID
- The number of rounds

```{r, warning = FALSE}
# Get start row and last PID
for (i in 1:length(s)){
   temp <- as.integer(str_extract_all(s[i],"^\\s+[1]\\s+"))
   temp2 <- as.integer(str_extract_all(s[i],"^\\s+[0-9]+"))
   if (is.na(temp) == FALSE){
     start_row <- i # Get the row that the actual games start on 
     cat("Start row is: ", start_row)
   }
   if (is.na(temp2) == FALSE){
     last_id <- temp2 # Get the PID of the last game
   }
}
cat("Last PID is: ",last_id) 

# Get the number of rounds
for (i in 1:length(s)){
  temp <- str_extract_all(s[i],"\\s+[0-9]+\\s+\\|\\s+$")
  temp <- str_split_fixed(temp, "\\|", n=Inf)
  if (temp[1][1] != "character(0)"){
    n_rounds <- as.integer(temp[1][1])
    cat("Number of rounds: ", n_rounds)
    break
  }                                 
}
```

Form one dataframe that has the following information:

- Player ID
- Player Name
- Total Number of Points
- State
- Players Pre-Rating

Form a matrix that has the PID followed by the opponent IDs, with NAs for no opponent for that round.

```{r simplify-s}

# Make the dataframes
p_info <- data.frame(
  p_id = integer(),
  p_name = character(),
  p_points = double(),
  p_state = character(),
  p_prerating = integer()
)

o_ids <- matrix(, nrow = last_id, ncol = n_rounds + 1)

for (i in seq(start_row,length(s),by =3)){
  
  ss_1 <- str_split_fixed(s[i],regex("|", literal=TRUE),n=Inf)
  # PID | P Name | Total Points | X Opponent PID | X Opponent PID | etc...
  # If X is W,L or D - read Opponent PID, increment total opponents
  p_id <- as.integer(ss_1[1])
  p_name <- str_trim(ss_1[2],side="both")
  p_points <- as.double(ss_1[3])
  
  ss_2 <- str_split_fixed(s[i+1],regex("|", literal=TRUE), n=Inf)
  # State | X/R: Player’s Pre-Rating
  p_state <- str_trim(ss_2[1],side="both")
  temp <- str_extract_all(ss_2[2],"[0-9]+")
  p_prerating <- as.integer(temp[[1]][2])
  
  start_i <- 4
  start_o <- 2
  o_ids[p_id, 1] <- p_id
  for (ii in start_i:(length(ss_1)-1)){
      o_ids[p_id, start_o] <- as.integer(str_extract(ss_1[ii],"[0-9]+"))
      start_o <- start_o + 1
  }
  
  p_temp <- data.frame(
    p_id,
    p_name,
    p_points,
    p_state,
    p_prerating)
  
  p_info <- rbind(p_info, p_temp)
}
```


Now that there is a dataframe with all the player information and a matrix with all of the player-opponent information, the calculation for the average pre-chess rating of opponents can be made for each player.

```{r}
pco_rating <- matrix(, nrow = last_id, ncol = 2)
for (i in 1:dim(o_ids)[1]){
  temp_sum <- 0
  temp_total <-0
  for (ii in 2:dim(o_ids)[2]){
    temp_o_id <- o_ids[i,ii]
    if (is.na(temp_o_id) == FALSE){
      temp_sum <- temp_sum + p_info$p_prerating[temp_o_id]
      temp_total <- temp_total + 1
    }
  }
  if (temp_total > 0){
    temp_average <- temp_sum/temp_total
  } else {
    temp_average <- NA # handle case where no opponents
  }
  pco_rating[i,1] <- o_ids[i,1]
  pco_rating[i,2] <- temp_average
}
```

Now, all relevant information is in the dataframe except the pre-chess rating of the opponents.
Do a quick check that the number of rows in the dataframe and the matrix match.

```{r}
dim(p_info)
dim(pco_rating)
```

Make a new dataframe that includes the rounded pre-chess rating of the opponents.

```{r}
df <- data.frame(
  p_name = character(),
  p_points = double(),
  p_state = character(),
  p_prerating = integer(),
  o_prerating = integer())

for (i in 1:dim(p_info)[1]){
  p_info$p_id[i]
  for (ii in 1:dim(pco_rating)[1]){
    if (p_info$p_id[i] == pco_rating[ii,1]){
      df_temp <-  data.frame( 
                    p_info$p_name[i],
                    p_info$p_points[i],
                    p_info$p_state[i],
                    p_info$p_prerating[i],
                    round(pco_rating[ii,2]))
      df <- rbind(df, df_temp)
      break
    }
  }
}
```

Do a quick check to see if the results make sense.

```{r}
head(df)
```

A brief check validates the data.

Write the dataframe to a csv file.

```{r}
write.csv(df, "chess_ratings.csv")
```

## Conclusion

This code successfully takes the input text file and produces an output file with the Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents.

This code is flexible in a few ways:

(1) The starting row can be in a different place, it is just the first row that starts with a `1`.
(2) The number of rounds can be varied, as it is detected based on the last number of the header row.
(3) The number of players can be varied, as the last players row is detected as the last row starting with a number.

For debugging purposes, rather than relying on the index of the `p_info` dataframe, the `o_ids` matrix, and the `pco_ratings` matrix to align and indicate the Player ID, I stored the Player ID in each. For `p_info` it is `p_info$p_id` and for `o_ids` and `pco_ratings` it is the first column. This way if an entry or entries are missing, the Player ID is still preserved and can be referenced back to. 

This code will include players in the output csv, even if the player did not have any opponents. In this case, the average pre-chess rating of opponents column would be `NA`. 

A shortcoming of this code is that it may not handle an incorrectly formatted row well. Depending on what the formatting error is, the code could get into a state that it would not recover from. 

To extend this code, I would take the input file and introduce different formatting errors to it and work on the error handling aspects of the code.